---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Allison Li"
date: "02142025"
output: github_document
---

### Email and name
```{r email and name}
library(usethis)
use_git_config(
  user.name = "Allison",
  user.email = "liy423@gmail.com"
)
```


### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox)
library(mosaicData) 
staff <- read_csv("data/instructional-staff.csv")
```

### Exercise 1

```{r line-plot}
staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))
staff_long

staff_long %>%
  ggplot(aes(x = year, y = value, color = faculty_type)) +
  geom_line()

staff_long %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type
  )) +
  geom_line() +
  labs(title = "Instructional staff employment trends",
       y = "percentage of hires",
       color = "faculty type")
```

According to the graph, we can clearly see that the Part-Time faculty employment trend has increased over the years while the Full-time Tenure-track Faculty and the Full-time Tenured Faculty showed clear decreasing trends. For Full-time Non-Tenured-Track Faculty, it seems that there are higher employment rate in 2011 comparing with 1975. For Graduate Student Employees, although there is a clear decrease in 1989, the rest employment rate remained relatively similar over the years, around 20%. 

### Exercise 2

If I want to show the trend of the part-time faculty employment rate comparing with other staff types, I would either make other types of faculty type different types of color grey, or I would make this line thicker so people would know what is the main point of my graph.   
Additionally, I believe we can also re-code the variables into part-time faculty and other types of faculty to highlight the comparison. 

```{r improved comparison visualization}
##I plan to use one line with color to represent part-time faculty employment rate trend while keep other lines grey to represent other instructional staff types.

staff_long %>%
  mutate(faculty_type_compare = if_else(faculty_type == "Part-Time Faculty", "Part-Time Faculty", "other")) %>% 
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type_compare
  )) +
  geom_line(size = 1) +
  scale_color_manual(values = c("Part-Time Faculty" = "#FF007F", "other" = "gray")) +
  labs(title = "Instructional staff employment trends comparison",
       y = "percentage of hires",
       color = "faculty type")
```

### Exercise 3

```{r refine visualization of fisheries}
##In my understanding, I believe one of the best ways to make a better visualization is to make a bar chart for the top 10 countries with the most fish caught and farmed.
fisheries <- read_csv("data/fisheries.csv")

##the top ten countries for farming fish
top10_aquaculture <- fisheries %>%
  arrange(desc(aquaculture)) %>%
  slice_head(n = 10)
##The top ten countries for fish caught
top10_capture <- fisheries %>%
  arrange(desc(capture)) %>%
  slice_head(n = 10)

##since the top ten countries for farming and capturing are different, I decided to make two bar chart with one for each. 
##Top 10 countries of capture fish visualization
top_countries_capture <- top10_capture %>%
  pivot_longer(cols = c(aquaculture, capture), names_to = "type", values_to = "production")

ggplot(top_countries_capture, aes(x = reorder(country, -production), y = production, fill = type)) +
  geom_bar(stat = "identity", position = "dodge") + 
  coord_flip() +
  labs(title = "Top 10 Countries in Fisheries Capture (Aquaculture vs. Capture)",
       x = "Country",
       y = "Production (metric tons)",
       fill = "Production Type")    

##Top 10 countries of farming fish visualization
top_countries_farm <- top10_aquaculture %>%
  pivot_longer(cols = c(aquaculture, capture), names_to = "type", values_to = "production")

ggplot(top_countries_farm, aes(x = reorder(country, -production), y = production, fill = type)) +
  geom_bar(stat = "identity", position = "dodge") + 
  coord_flip() +
  labs(title = "Top 10 Countries in Fisheries Farm (Aquaculture vs. Capture)",
       x = "Country",
       y = "Production (tons)",
       fill = "Production Type")  
```

### Exercise 4

```{r smokers in Whickham}
###install.packages("mosaicData")
###install.packages("performance")
library(tidyverse)
library(mosaicData)

data(Whickham)
?Whickham
```

4.1. According to the codebook, I think the data comes from observational data. Because it is simply asking questions including the age of the participant, whether the participant is dead of alive (outcome), and whether the participant smoke or not. These questions are not manipulated or controled, so this is an observation.
4.2. There are 1314 observations in the dataset, with each of them representing the basic information for each participant about whether they smoke or not, age, and the living outcomes. 
4.3. There are three variables. The outcome variable is categorical variable; The smoker variable is categorical variable; The age variable is numeric variable. 
4.4.
4.5. 

### Exercise 5
